Toward a clearer understanding of the inverse association between cancer and dementia

### 2021-04-27 06_presentations/ser2021/index.Rmd
ser presentation updates

```{r setup, include=FALSE, }
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(collapse = TRUE, fig.retina = 3, warning = FALSE,
                      message = FALSE, echo = FALSE, fig.showtext = TRUE, cache = TRUE)
library(WeightIt)
library(cobalt)
library(splines)
library(knitr)
library(kableExtra)

@@

## Aims

To conceptualize and progressively build a causal directed acyclic graph to connect this particular causal question to the observable data and the assumptions we rely on to identify the effect.

---
### Pin1-targeting drug

.center[
```{tikz, fig.ext = 'png', echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$D$};
\node [right of = p] (y) {$Y_{t}$};
\draw[->] (p) -- (y);
\end{tikzpicture}
```

What is the effect of a Pin1 targeting drug $(D)$ in the risk of ADRD over time *t* $(Y_{t})$  compared to standard treatments?

@@

---
### Pin1 expression as biomarker

.center[
```{tikz, fig.ext = 'png', echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$P_0$};
\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
\node [right of = p] (y) {$Y_{t}$};
\draw[->] (l) -- (p);
\draw[->, violet] (p) -- (y);
\draw[->] (l) -- (y);
\end{tikzpicture}
```

$P_0$ = Pin1 over-expression; $L$ = shared factors between $P$ and $Y_t$

@@

---
### Cancer diagnosis as the proxy of Pin1 biomarker

.center[
```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = ""}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$P_0$};
\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
\node [above of = p, yshift = -1cm](pstar) {$P^*$};
\node [right of = p] (y) {$Y_{t}$};
\draw[->, violet] (p) -- (pstar);
\draw[->, violet] (p) -- (y);
\draw[->] (l) -- (p);
\draw[->] (l) -- (y);

\end{tikzpicture}
```

$P_0$ = Pin1 over-expression; $P^*$ = Cancer diagnosis; $L$ = shared factors between $P$ and $Y_t$ = ADRD diagnosis
]

@@

---
### Zoom in the P and P* association

.center[
```{tikz, fig.ext = 'png', echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$P_0$};
\node [right of = p] (d) {$D_{k} = 0$};
\node [right of = d] (pstar) {$P^*_{k+1}$};
\draw[->] (p) -- (d);
\draw[->, violet] (p) to [out=45,in=135, looseness=0.4] (pstar);
\draw[->] (d) -- (pstar);
\end{tikzpicture}
```

<br> <br>

$P_0$ = Pin1 over-expression; $D_{k} = 0$ = surviving by time $k$; $P^*_{k+1}$ = Cancer diagnosis by time $k+1$ ;

???

---
### Selection bias in the P and P* association

.center[
```{tikz, fig.ext = 'png', echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$P_0$};
\node [right of = p] (d) {$D_{k} = 0$};
\node [right of = d] (pstar) {$P^*_{k+1}$};
\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$C_1$};
\node [below of = d, yshift = 1cm, xshift = 1cm](c) {$C_2$};
\draw[->] (p) -- (d);
\draw[->, violet] (p) to [out=45,in=135, looseness=0.4] (pstar);
\draw[->] (d) -- (pstar);
\draw[->] (l) -- (p);
\draw[->] (l) -- (d);
\draw[->] (c) -- (pstar);
\draw[->] (c) -- (d);
\end{tikzpicture}
```

<br> <br>

$P_0$ = Pin1 over-expression; $D_{k} = 0$ = surviving by time $k$; $P^*_{k+1}$ = Cancer diagnosis by time $k+1$; $C_1$, $C_2$ shared factors such as smoking

???
In some way we need to assume that those who died would have the same risk of having a cancer diagnosis had they not died 
---
### Direct effect vs. total effect, potential bias

.center[
```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = ""}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$P$};
\node [above of = p, yshift = -1cm](pstar) {$P^*$};
\node [right of = p] (d) {$D_{t}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = d, yshift = 1cm, xshift = +1cm](c) {$C$};
\draw[->, violet] (p) -- (pstar);
\draw[->] (p) -- (d);
\draw[->, very thick] (d) -- (y);
\draw[->] (c) -- (d);
\draw[->] (c) -- (y);
\draw[->, violet] (p) to [out=45,in=135, looseness=0.4] (y);
\end{tikzpicture}
```
]

???
In this setting death is treated as a censoring event and it can be interpreted as those who died would have the same risk of developing dementia if prevented from dying than those who remained alive and free of dementia at a given time point, conditional on shared causes of dementia and death

---
### All pieces together

.center[
```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = ""}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$P$};
\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
\node [right of = p] (dp) {$D_t$};
\node [right of = dp] (pstar) {$P*_t$};
\node [right of = pstar] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [above of = dp, yshift = -1cm, xshift = +1cm](c) {$C_1$};
\node [above of = d, yshift = -1cm, xshift = +1cm](cd) {$C_2$};
\draw[->] (p) -- (dp);
\draw[->] (dp) -- (pstar);
\draw[->, violet] (p) to [out=-45,in=-135, looseness=0.3] (pstar);
\draw[->] (pstar) -- (d);
\draw[->] (p) to [out=-45,in=-135, looseness=0.3] (d);
\draw[->, violet] (p) to [out=-45,in=-135, looseness=0.3] (y);
\draw[->] (d) -- (y);
\draw[->] (l) -- (p);
\draw[->] (l) to [out=0,in=270, looseness=0.4] (y);
\draw[->] (c) -- (dp);
\draw[->] (c) -- (pstar);
\draw[->] (cd) -- (d);
\draw[->] (cd) -- (y);
\end{tikzpicture}
```

At best we can test the sharp null hypothesis
]

@@

- Participants between 60 and 70 years old at study entry.

@@

Three definitions of the proxy for Pin1 over-expression:

- _"Cancer ever vs. never"_

- Cancer as a _"time-varying"_

- _"Time to cancer diagnosis"_

### Methods

- Inverse probability treatment weights to address confounding

  + *Age, sex, education, smoking status*

@@

- Inverse probability censoring weights for death to relax the independent censoring assumption

@@

- We estimate the risk of dementia using a weighted Kaplan-Meier estimator.

### Methods

Since the independent censoring assumption is untestable, we can consider extreme scenarios of independence and fit Peterson upper and lower bounds.

1) Total independence, that refers to an scenario were those who died would never develop dementia (lower bound).
  + Aalen-Johansen estimator treating death as a competing event.

2) Complete dependency, that refers to an scenario where those who died would have dementia prior to death (upper bound). 
  + The Kaplan Meier estimator for the combined outcome of dementia or death.

@@

.pull-left[
```{r, echo = FALSE}
plot_distribution +
    theme_xaringan()
```
]

.pull-right[
- 24% had cancer, the median age of cancer diagnosis was 73 (IQR: 69-77).

- Among participants with cancer, 6% had dementia and 63% (n = 549) died.

- Among participants free of cancer diagnosis, 14% had dementia and 23% died.
]
---
### Cancer as: never vs. ever

@@

### Time to cancer

.pull-left[
```{r}
plot_3a
```

@@

.pull-right[
```{r}
plot_3b
```

@@

.center[
```{r, echo = FALSE}
tb <- table_results %>%
  filter(model != "Unadjusted") %>% 
  select(-c(3, 4)) %>%
  rename(RD = rd, RR = rr, `HR (95%CI)` = hr)

>>>>

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE, fig.showtext = TRUE, cache = TRUE)

library(splines)
library(knitr)
library(kableExtra)

@@

### Pin1-targeting drug

```{tikz, echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
\node [right of = p, gray!30] (dp) {$D_t$};
\node [right of = dp, gray!30] (pstar) {$P^*_t$};
\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
\draw[->, gray!30] (p) -- (dp);
\draw[->, gray!30] (l) -- (p);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->, gray!30] (dp) -- (pstar);
\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (pstar);
\draw[->, gray!30] (pstar) -- (d);
\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
\draw[->, gray!30] (d) -- (y);
\draw[->, gray!30] (c) -- (dp);
\draw[->, gray!30] (c) -- (pstar);
\draw[->, gray!30] (cd) -- (d);
\draw[->, gray!30] (cd) -- (y);
\end{tikzpicture}
```

$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis

@@

---
### Pin1 expression as biomarker (Problem 1: _Confounding_)

```{tikz, echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -1cm, xshift = 1cm](l) {$L$};
\node [right of = p, gray!30] (dp) {$D_t$};
\node [right of = dp, gray!30] (pstar) {$P^*_t$};
\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
\draw[->, gray!30] (p) -- (dp);
\draw[->] (l) -- (p);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, semithick] (p) to [out=45,in=135, looseness=0.3] (y);
\draw[->, gray!30] (dp) -- (pstar);
\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar);
\draw[->, gray!30] (pstar) -- (d);
\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (d);
\draw[->, gray!30] (d) -- (y);
\draw[->, gray!30] (c) -- (dp);
\draw[->, gray!30] (c) -- (pstar);
\draw[->, gray!30] (cd) -- (d);
\draw[->, gray!30] (cd) -- (y);
\end{tikzpicture}
```

$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $L$ = Shared causes of $P_0$ and $Y_{t+1}$

@@

---
### Cancer diagnosis as the proxy of Pin1 biomarker (Problem 2: _Time zero_)

```{tikz, echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
\node [right of = p, gray!30] (dp) {$D_t$};
\node [right of = dp] (pstar) {$P^*_t$};
\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
\draw[->, gray!30] (p) -- (dp);
\draw[->, gray!30] (l) -- (p);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->, gray!30] (dp) -- (pstar);
\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
\draw[->, gray!30] (pstar) -- (d);
\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
\draw[->, gray!30] (d) -- (y);
\draw[->, gray!30] (c) -- (dp);
\draw[->, gray!30] (c) -- (pstar);
\draw[->, gray!30] (cd) -- (d);
\draw[->, gray!30] (cd) -- (y);
\end{tikzpicture}
```

$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $P^*_t$ = Cancer diagnosis

@@

---
### Death prior to cancer diagnosis (Problem 3: _Survival bias_)

```{tikz, echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
\node [right of = p] (dp) {$D_t$};
\node [right of = dp] (pstar) {$P^*_t$};
\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
\draw[->] (p) -- (dp);
\draw[->, gray!30] (l) -- (p);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->] (dp) -- (pstar);
\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
\draw[->, gray!30] (pstar) -- (d);
\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
\draw[->, gray!30] (d) -- (y);
\draw[->, gray!30] (c) -- (dp);
\draw[->, gray!30] (c) -- (pstar);
\draw[->, gray!30] (cd) -- (d);
\draw[->, gray!30] (cd) -- (y);
\end{tikzpicture}
```

$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $P^*_t$ = Cancer diagnosis; $D_t$ = Death prior to cancer diagnosis
???

---
### Death prior to cancer diagnosis (Problem 3: _Survival bias_)

```{tikz, echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
\node [right of = p] (dp) {$D_t$};
\node [right of = dp] (pstar) {$P^*_t$};
\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = dp, yshift = 1cm, xshift = 1cm](c) {$C_1$};
\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
\draw[->] (p) -- (dp);
\draw[->, gray!30] (l) -- (p);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->] (dp) -- (pstar);
\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
\draw[->, gray!30] (pstar) -- (d);
\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
\draw[->, gray!30] (d) -- (y);
\draw[->] (c) -- (dp);
\draw[->] (c) -- (pstar);
\draw[->, gray!30] (cd) -- (d);
\draw[->, gray!30] (cd) -- (y);
\end{tikzpicture}
```

$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $P^*_t$ = Cancer diagnosis; $D_t$ = Death prior to cancer diagnosis; $C_1$ = Shared causes of $D_t$ and $P^*$
???

---
### Death prior to dementia diagnosis (Problem 3: _Survival bias_)

```{tikz, echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
\node [right of = p, gray!30, gray!30] (dp) {$D_t$};
\node [right of = dp, gray!30] (pstar) {$P^*_t$};
\node [right of = pstar] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
\draw[->, gray!30] (p) -- (dp);
\draw[->, gray!30] (l) -- (p);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->, gray!30] (dp) -- (pstar);
\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar);
\draw[->, gray!30] (pstar) -- (d);
\draw[->] (p) to [out=+45,in=+135, looseness=0.3] (d);
\draw[->] (d) -- (y);
\draw[->, gray!30] (c) -- (dp);
\draw[->, gray!30] (c) -- (pstar);
\draw[->, gray!30] (cd) -- (d);
\draw[->, gray!30] (cd) -- (y);
\end{tikzpicture}
```

$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $D_{t+1}$ = Death prior to dementia diagnosis

---
### Death prior to dementia diagnosis (Problem 3: _Survival bias_)

```{tikz, echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
\node [right of = p, gray!30, gray!30] (dp) {$D_t$};
\node [right of = dp, gray!30] (pstar) {$P^*_t$};
\node [right of = pstar] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
\node [below of = d, yshift = 1cm, xshift = 1cm](cd) {$C_2$};
\draw[->, gray!30] (p) -- (dp);
\draw[->, gray!30] (l) -- (p);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->, gray!30] (dp) -- (pstar);
\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar);
\draw[->, gray!30] (pstar) -- (d);
\draw[->] (p) to [out=+45,in=+135, looseness=0.3] (d);
\draw[->] (d) -- (y);
\draw[->, gray!30] (c) -- (dp);
\draw[->, gray!30] (c) -- (pstar);
\draw[->] (cd) -- (d);
\draw[->] (cd) -- (y);
\end{tikzpicture}
```

$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $D_{t+1}$ = Death prior to dementia diagnosis; $C_2$ as shared causes of $D_{t+1}$ and $Y_{t+1}$

@@

- Participants between 60 and 70 years old at study entry, free of cancer and dementia diagnosis at baseline

@@

@@

- Cancer as _"time-fixed ever vs. never"_ vs. cancer as a _time-varying proxy_

- Inverse probability treatment weights to address confounding (IPTW)

  + *Age, sex, Apoe4, education*

- Inverse probability censoring weights for death (IPCW)

@@

- Risk of dementia is estimated with a weighted Kaplan-Meier estimator.

---
### Peterson upper and lower bounds

- **1) Lower Bound**: Total independence, that refers to an scenario were those who died would never develop dementia (Death as competing event)
--

- **2) Upper Bound**: Complete dependency, that refers to an scenario where those who died would have dementia prior to death (Composite outcome, death + dementia)

@@

.center[
```{r, echo = FALSE, fig.retina = 3}
plot_distribution +
    theme_xaringan()
```

In the cancer arm 6% had dementia and 63% died, in the free of cancer arm 14% had dementia and 23% died.

---
### Cancer as time-fixed ever vs. never proxy

@@

### Cancer diagnosis as time-varying proxy

.pull-left[
```{r}
plot_2b
```

@@

.pull-right[
```{r}
plot_2c
```

@@

.center[
```{r, echo = FALSE}
tb <- table_results %>%
  filter(model != "Unadjusted") %>%
  select(-c(3, 4, 5)) %>%
  rename(RR = rr, `HR (95%CI)` = hr) %>% 
  filter(Proxy != "Time to cancer")
